To Airbnb?: Factors impacting short-term leasing preference

### 2018-11-14 papers/full_paper.Rmd
Final edits for submission #3

The growth of Airbnb and other short-term rental platforms have presented absentee owners of urban residential properties with a choice of leasing strategy: traditional long-term rental or a short-term approach ("Airbnb-ing").  In this paper we identify those situations -- location, structure type and property characteristics -- that lead to the highest likelihood of favoring a short-term strategy over a long term one. Additionally, tests on the impacts of hosting policies suggest that even the right property may need the right owner(s) or strategy to make short-term rental the more profitable approach.

@@

We adopt a scenario-based approach that compares observed annual revenues for short-term rental (STR) properties with an estimate of their expected long-term rental revenues, netting out costs that can be estimated or assumed with some generality. We handle potential undercounting of costs through sensitivity tests that follow our initial assessment.

Our results show that the feasibility of a greater revenues through a short-term rental -- what we have termed ''Short-Term Preference'' -- is dependent on property type, product, location and host policies.  In short, the answer to whether or not a owner of an urban residential property is likely to earn more through a short-term leasing strategy than a long-term one is ''it depends''.

@@

Existing research into short-term rental markets is sparse. Within the existing work, the focus remains centered on rural and/or amenity-based STRs such as beach-front and ski-resort access properties.  This focus results from the fact that, until a few years ago, the STR market itself operated almost wholly within these market segments. The shift to small, urban dwellings within the STR market is relatively new and explains the dearth of past work in this direction. With the rapid growth of Airbnb, and the controversies surrounding it, interest in the short-term rental market has increased among research as well as investors and industry analysts.

@@

A critical theme that arises out of the extant literature is that market fundamentals for primary dwellings (owner-occupied and long-term rentals) may differ from those of vacation homes (or, by association, short-term rentals). Or, in other words, the factors that influence prices and rents in the two are not equivalent. @cho2003impacts find that the second home market is more sensitive to proximity to environmental amenities than the owner occupied market. In Switzerland, @soguel2008impact test to see if second homes pay a higher premium for ski resort access.  Their results fail to show a statistically significant difference in the impact of ski resorts between markets, but do find that other property characteristics have heterogenous effects between the two markets. In examining the difference between the urban and rural STR markets,  @nilsson2015influence makes a key distinction finding that rural vacation home markets value amenities whereas urban holiday properties tend to derive large premiums from proximity to attractive residential areas.

@@

Unfortunately, we cannot not directly observe properties that are utilized in both a short- and long-term leasing strategy. Therefore, comparisons are done with actual short-term but estimated or imputed long-term revenues. Initially, we also attempted to estimate short-term rental revenue for long-term rental properties but found the process of imputing occupancy rates and nightly rate structures too inaccurate to provide valid comparisons.

@@

where $Rate_{N}$ is the average nightly rate and $O_{Total}$ are the total number of occupancies (bookings) over the 12-month period. For long-term rentals the net revenue, $Rev_{LT}$, is defined as:

@@

Data for this study were gathered from two different sources, one for the short-term rental market and one for the long-term rental market.  Short-term rental market observations are from the Airbnb online portal. Airbnb is the largest short-term rental hub in the Melbourne region. As opposed to similar sites such as VRBO.com and Homeaway.com which focus on vacation homes in tourist dominated areas such as beach towns and ski resorts, Airbnb offers considerable coverage in urban areas and, therefore, directly competes to a greater degree with more traditional long-term rentals than the other sites.^[In fact, our choice of study area was done so to specifically minimize the number of potential traditional vacation homes in our sample.] The data on the Airbnb market in Melbourne have been purchased from www.airdna.co, a data provider specializing in Airbnb data collection and analysis. The data offers both nightly observations indicating the advertised price, the occupancy status and a unique reservation identifier as well as general property and host policy information.

@@

Data on the long-term rental market was provided by Australia Property Monitors (APM), a Fairfax Group company. The long-term lease contains both property specific information (beds, baths, location, etc.) as well as a history of listing prices.  These data cover the period from January 2014 to December 2015.

@@

Table 1 highlights the filters applied to the data.  Filters were applied in the order listed in the table.  Details on two filters -- property extent, host classification and location -- are described more fully below, more information on the other filter can be found in the online code repository for this research (see the Reproducibility section at the end of this paper).  After employing all filters we were left with 2,415 short term and 64,305 long term observations.

@@

We begin by eliminating any observations in the outer suburbs.  Very few short-term rentals were located in these areas and those that were have a much higher likelihood of being vacation homes; a group of properties uses that doesn't match well to our research question. While the Inner and Middle suburb designation is a useful starting place, we belief that a few additional spatial features can also influence the short-term market.  While there are many possible additional features, we consider two to be the most likely to influence rates or occupancy: 1) Proximity to beaches; and 2) Proximity to key tourist activities and events.

@@

Second, many of the short-term properties have a significant number of blocked days (up to 255), days in which the property was not offered on Airbnb.  Here too, there is a problem with comparing observed revenues from a property with a 20\% block rate with a hypothetical year long revenue stream, especially if the property was leased out on other portals during those blocked days.  And, even if we knew that the blocked days were simply due to the host wanting to use the property themselves during that time, this deduction in revenue should factor into the comparative analyses we make.  In short, in either case the actual observed short-term revenue would be biased on the low end compared to what a revenue maximizing owner would be likely able to accomplish.

@@

Overall, 57.3%^[If we just look at actual revenue, ignoring blocked days and properties with less than a full 12 months of listing, the figure is 42.5%] of the 2,415 short-term urban rental properties in our sample were more profitable as a short-term rental than it would have been if rented long-term over the course of the year. While illustrative, a simple univariate analysis does not fully control for location, property characteristics or owner actions -- which are a key component of short-term rental success.  To better understand which factors are related to a higher likelihood of short-term preference, we develop a set of three increasingly complex, logistic regression models.

@@

In Model 1, we see that when controlling for structural characteristics, houses have a significantly lower probability of being more profitable in short-term leasing than apartments. Additionally, 1 bed / 1 bath units (reference category) show greater likelihood of short-term preference than 2 bed / 2 bath combinations. This could be due to the fact that 2 bed / 2 bath units may be very attractive in the long-term market as they are a useful product for two, non-family roomates to share. Larger units, 3 bedrooms plus, generally show no statistical difference in preference probability to the 1 bed / 1 bath reference unit. In other words, larger units are equally desirable to tourists and local residents alike.

@@

Finally, in Model 3 we add in host policies. Allowing a greater number of guests per bedroom results in a higher probability of short-term preference.  This variable may be operating as proxy for additional sleeping spaces which allow for a higher nightly rate, ceteris paribus, to be charged.  The greater the minimum stay, the lower the probability of short-term preference.  Longer minimum stays may equate to lower occupancy rates as one or two night travelers are prevented from booking.  Owners with longer minimum stays may be looking to minimize costs such as cleaning and administration.  Finally, flexible cancellation policies result in a considerably lower probability of short-term preference. Strict cancellation may be a luxury available only for the best and most highly desired properties on the market and, within our model, may be acting as a proxy for some unobserved measure of quality or demand while a flexible cancellation may be signalling the opposite.

@@

Although it gathers considerable interest in both economic and public policy circles, analyses of short-term markets and their comparison to more traditional long-term leasing strategies remains relatively unaddressed. The Airbnb market is Melbourne is rapidly expanding as total listings on the portal have jumped from around 7,000 in 2014 to more than 20,000 in August 2016.  However, of this total only 7,400 or so of the properties are entire homes or apartments that were listed as available for at least one day during the September 1, 2015 to August 31, 2016 time frame.  Of these, only 2,415 are common bed/bath configurations with non-outlying rates and were offered for short term lease at least 75% of the year. Overall, the Airbnb market is very small relative to the long-term market. We caution future researchers and policy analysts to look deeply at initial estimates of Airbnb counts as many of the listings on the market are either room shares and/or not active.

@@

The authors would like to thank the Real Estate Research Institute (RERI) for funding this work and the Australian Property Monitors (APM) for providing the long-term rental data. We'd also like to thank an anonymous reviewer for suggesting the matched pairs approach as a check on our hedonic model.

>>>>

The growth of Airbnb and other short-term rental platforms have presented absentee owners of urban residential properties with a choice of leasing strategy: traditional long-term rental or a short-term approach ("Airbnb-ing").  In this paper we identify those situations -- location, structure type and property characteristics -- that lead to the highest likelihood of favoring a short-term strategy over a long term one. Additionally, we test the impacts of hosting policies the results of which suggest that even the right property may need the right owner(s) or strategy to make short-term rental the more profitable approach.

@@

We adopt a scenario-based approach that compares observed annual revenues for short-term rental (STR) properties with an estimate of their expected long-term rental revenues, netting out costs that can be estimated or assumed based on property, and not owner, characteristics. We handle potential undercounting of costs or over-estimation of long-term revenues through sensitivity tests that follow our initial assessment.

Our results show that the feasibility of a greater revenues through a short-term rental -- what we have termed ''Short-Term Preference'' -- is dependent on property type, product, location and host policies.  In short, the answer to whether or not the owner of an urban residential property is likely to earn more through a short-term leasing strategy than a long-term one is ''it depends''.

@@

Existing research into short-term rental markets is sparse. Within the existing work, the focus remains centered on rural and/or amenity-based STRs such as beach-front and ski-resort access properties.  This focus results from the fact that, until a few years ago, the STR market itself operated almost wholly within these market segments. The shift to small, urban dwellings within the STR market is relatively new and explains the dearth of past work in this direction. With the rapid growth of Airbnb, and the controversies surrounding it, interest in the short-term rental market has increased among researchers as well as investors and industry analysts.

@@

A critical theme that arises out of the extant literature is that market fundamentals for primary dwellings (owner-occupied and long-term rentals) may differ from those of vacation homes (or, by association, short-term rentals). Or, in other words, the factors that influence prices and rents in the two are not equivalent. @cho2003impacts find that the second home market is more sensitive to proximity to environmental amenities than the owner occupied market. In Switzerland, @soguel2008impact test to see if second homes pay a higher premium for ski resort access.  Their results fail to show a statistically significant difference in the impact of ski resorts between markets, but do find that other property characteristics have heterogenous effects between the two markets. Finally, in examining the difference between the urban and rural STR markets,  @nilsson2015influence makes a key distinction finding that rural vacation home markets value amenities whereas urban holiday properties tend to derive large premiums from proximity to attractive residential areas.

@@

Unfortunately, we cannot directly observe properties that are utilized in both a short- and long-term leasing strategy. Therefore, comparisons are done with actual short-term but estimated or imputed long-term revenues. Initially, we also attempted to estimate short-term rental revenue for long-term rental properties but found the process of imputing occupancy rates and nightly rate structures too inaccurate to provide valid comparisons.

@@

where $Rate_{N}$ is the median nightly rate and $O_{Total}$ are the total number of occupancies (bookings) over the 12-month period. For long-term rentals the net revenue, $Rev_{LT}$, is defined as:

@@

Data for this study were gathered from two different sources, one for the short-term rental market and one for the long-term rental market.  Short-term rental market observations are from the Airbnb online portal. Airbnb is the largest short-term rental hub in the Melbourne region. As opposed to similar sites such as VRBO.com and Homeaway.com which focus on vacation homes in tourist dominated areas such as beach towns and ski resorts, Airbnb offers considerable coverage in urban areas and, therefore, directly competes to a greater degree with more traditional long-term rentals than the other sites.^[In fact, our choice of study area was done so to specifically minimize the number of potential traditional vacation homes in our sample.] The data on the Airbnb market in Melbourne have been purchased from www.airdna.co, a data provider specializing in Airbnb data collection and analysis. The data offer both nightly observations indicating the advertised price, the occupancy status and a unique reservation identifier as well as general property and host policy information.

@@

Data on the long-term rental market were provided by Australia Property Monitors (APM), a Fairfax Group company. The long-term lease contains both property specific information (beds, baths, location, etc.) as well as a history of listing prices.  These data cover the period from January 2014 to December 2015.

@@

Table 1 highlights the filters applied to the data.  Filters were applied in the order listed in the table.  Details on three of the filters -- property extent, host classification and location -- are described more fully below, more information on the other filter can be found in the online code repository for this research (see the Reproducibility section at the end of this paper).  After employing all filters we were left with 2,415 short term and 64,305 long term observations.

@@

We begin by eliminating any observations in the outer suburbs.  Very few short-term rentals were located in these areas and those that were have a much higher likelihood of being vacation homes; a group of properties uses that doesn't match well to our research question. While the Inner and Middle suburb designation is a useful starting place, we believe that a few additional spatial features can also influence the short-term market.  While there are many possible additional features, we consider two to be the most likely to influence rates or occupancy: 1) Proximity to beaches; and 2) Proximity to key tourist activities and events.

@@

Second, many of the short-term properties have a significant number of blocked days (up to 91), days in which the property was not offered on Airbnb.  Here too, there is a problem with comparing observed revenues from a property with a 20\% block rate with a hypothetical year long revenue stream, especially if the property was leased out on other portals during those blocked days.  And, even if we knew that the blocked days were simply due to the host wanting to use the property themselves during that time, this deduction in revenue should factor into the comparative analyses we make.  In short, in either case the actual observed short-term revenue would be biased on the low end compared to what a revenue maximizing owner would be likely able to accomplish.

@@

Overall, 57.3%^[If we just look at actual revenue, ignoring blocked days and properties with less than a full 12 months of listing, the figure is 42.5%] of the 2,415 short-term urban rental properties in our sample were more profitable as a short-term rental than it would have been if rented long-term over the course of the year. While illustrative, a simple univariate analysis does not fully control for location, property characteristics or owner actions -- all of which are a key component of short-term rental success.  To better understand which factors are related to a higher likelihood of short-term preference, we develop a set of three increasingly complex, logistic regression models.

@@

In Model 1, we see that when controlling for structural characteristics, houses have a significantly lower probability of being more profitable in short-term leasing than apartments. Additionally, 1 bed / 1 bath units (reference category) show greater likelihood of short-term preference than 2 bed / 2 bath combinations. This could be due to the fact that 2 bed / 2 bath units may be very attractive in the long-term market as they are a useful product for two, non-family roommates to share. Larger units, 3 bedrooms plus, generally show no statistical difference in preference probability to the 1 bed / 1 bath reference unit. In other words, larger units are equally desirable to tourists and local residents alike.

@@

Finally, in Model 3 we add in host policies. Allowing a greater number of guests per bedroom results in a higher probability of short-term preference.  This variable may be operating as proxy for additional sleeping spaces which allow for a higher nightly rate, ceteris paribus, to be charged.  The greater the required minimum stay, the lower the probability of short-term preference.  Longer minimum stays may equate to lower occupancy rates as one or two night travelers are prevented from booking.  Finally, flexible cancellation policies result in a considerably lower probability of short-term preference. Strict cancellation may be a luxury available only for the best and most highly desired properties on the market and, within our model, may be acting as a proxy for some unobserved measure of quality or demand while a flexible cancellation may be signalling the opposite.

@@

Although it gathers considerable interest in both economic and public policy circles, analyses of short-term markets and their comparison to more traditional long-term leasing strategies remains relatively unaddressed. The Airbnb market in Melbourne is rapidly expanding as total listings on the portal have jumped from around 7,000 in 2014 to more than 20,000 in August 2016.  However, of this total only 7,400 or so of the properties are entire homes or apartments that were listed as available for at least one day during the September 1, 2015 to August 31, 2016 time frame.  Of these, only 2,415 are common bed/bath configurations with non-outlying rates and were offered for short term lease at least 75% of the year. Overall, the Airbnb market is very small relative to the long-term market. We caution future researchers and policy analysts to look deeply at initial estimates of Airbnb counts as many of the listings on the market are either room shares and/or not active.

@@

The authors would like to thank the Real Estate Research Institute (RERI) for funding this work and the Australian Property Monitors (APM) for providing the long-term rental data. We'd also like to thank an anonymous reviewer for suggesting the matched pairs approach as a check on our hedonic model and another for helping to shape our reseearch question.
